---
title: "PS5 Template"
author: "Arushi Sharma"
date: "2/13/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Disclaimer: this serves as a template for you to compose your answer for PS5. It is 100% hardcoded, so error may exists. Please refer to the official spec for a full list of questions and requirements.

# Instructions
This is a problem set about rmarkdown and plotting (using ggplot). Unlike the previous problem sets, this one does not give you a ready-made GH repo with a code file–it is now your task to create a repo and include your rmarkdown file in there.\

You should answer the questions below in that file, knit it, and submit both the compiled html and link to your repo on canvas.\

- This problem sets asks you to write extensively when commenting your results. Please write clearly! Answer questions in a way that if the code chunks are hidden then the result is still readable! \
    * All substantial questions need explanations. You do not have to explain the simple things like “how many rows are there in data”, but if you make a plot of life expectancy, then you should explain what does the plot tell you. \
    * Write explanations as markdown and use the styles like bold and italic as appropriate.\
- Do not print too much results. It is all well to print a few lines of data for evaluation/demonstration purposes. But do not print dozens (or thousands!) of lines–no one bothers to look at that many numbers. You will lose points for annoying others (here your graders, but later potentially your boss). \
- Do not make code lines too long. 80-100 characters is a good choice. Your grader may not be able to follow all the code if the line is too long–most of us are using small laptop screens! (And again–you want to keep your graders happy!) \

# Gapminder data
We use gapminder dataset, downloaded from [**https://www.gapminder.org/data/**](https://www.gapminder.org/data/), however, the data structure there is quire complex, please use the dataset provided on canvas (in files/data). \
The variables are: \
**name** country name \
**iso3** 3-letter country code \
**iso2** 2-letter country code \
**region** broad geographic region \
**sub-region** more precise region \
**intermediate-region** \
**time** year \
**totalPopulation** total population \
**GDP_PC** GDP per capita (constant 2010 US$) \
**accessElectricity** Access to electricity (% of population) \
**agriculturalLand** Agricultural land (sq. km) \
**agricultureTractors** Agricultural machinery, tractors (count) \
**cerealProduction** Cereal production (metric tons) \
**feritilizerHa** Fertilizer consumption (kilograms per hectare of arable land) \
**fertilityRate** total fertility rate (births per woman) \
**lifeExpectancy** Life expectancy at birth, total (years) \
**childMortality** Mortality rate, under-5 (per 1,000 live births) \
**youthFemaleLiteracy** Literacy rate, youth female (% of females ages 15-24) \
**youthMaleLiteracy** Literacy rate, youth male (% of males ages 15-24) \
**adultLiteracy** Literacy rate, adult total (% of people ages 15 and above) \
**co2** CO2 emissions (kt) \
**greenhouseGases** Total greenhouse gas emissions (kt of CO2 equivalent) \
**co2_PC** CO2 emissions (metric tons per capita) \
**pm2.5_35** PM2.5 pollution, population exposed to levels exceeding WHO Interim Target-1 value 36ug/m3 \
**battleDeaths** Battle-related deaths (number of people) \

# 1 Load and check data (5pt)
You first task is to do a very simple data check: 

1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See [**https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating**](https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating). \

```{r, message=FALSE}
library(tidyverse)
```

2. (2pt) Load data. How many rows/columns do we have?  \
```{r}
gm <- read_delim("data/gapminder.csv.bz2")
```
This dataset contains `r nrow(gm)` rows of observations and `r ncol(gm)` columns.

3. (2pt) Print a small sample of data. Does it look OK? 
```{r}
gm %>% 
  sample_n(5)
```
  
---------------------------------------------------------------------


# 2 Descriptive statistics (15pt)

1. (3pt) How many countries are there in the dataset? Analyze all three: *iso3*, *iso2* and *name*.
```{r}
gm %>% 
  summarise(uniques = n_distinct(iso2))

gm %>% 
  summarise(uniques = n_distinct(iso3))

gm %>% 
  summarise(uniques = n_distinct(name))
```

2. If you did this correctly, you saw that there are more iso-2 codes than names, and there are even more *iso3*-codes. What is going on? Can you find it out?
`r nrow(gm)`

    (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
```{r}
gm %>% 
group_by(iso2) %>% 
  summarise(uniques= n_distinct(name)) %>% 
  arrange(desc(uniques))

```


```{r}
gm %>% 
filter(is.na(iso2)) %>% 
  distinct(name)
```


the iso2 code of NA corresponds to both an NA value name and also the country of Namibia.

   
    (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are *CHANISL* and *NLD CURACAO*.
```{r}
gm %>% 
group_by(name) %>% 
  summarise(uniques= n_distinct(iso3)) %>% 
  arrange(desc(uniques))

```

```{r}
gm %>% 
filter(is.na(name)) %>% 
  distinct(iso3)
```
These countries are CHANISL, GBM, KOS, and NLD_CURACAO.


3. (2pt) What is the minimum and maximum year in these data?

```{r}
gm %>% 
 summarise(min(time,na.rm=TRUE),max(time,na.rm=TRUE))
```


---------------------------------------------------------------------

# 3 CO2 emissions (30pt)
Next, let’s analyze CO2 emissions.

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing *CO2* and *co2_PC*. Which years have most missing data? \
```{r}
gm %>% 
  group_by(time) %>% 
  summarise(missing=sum(is.na(co2))) %>% 
  arrange(desc(missing)) %>% 
  head(5)
```



```{r}
gm %>% 
  group_by(time) %>% 
  summarise(missing=sum(is.na(co2_PC))) %>% 
  arrange(desc(missing)) %>% 
  head(5)

```


The years 2017, 2018, and 2019 have the most missing data.


2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see. 
```{r}
gm %>% 
filter( iso2 %in% c("IN","CN","US","SE","VN") ) %>% 
filter(!is.na(co2)) %>% 
ggplot(aes(x=time, y=co2, col=name))+geom_line()+

labs(x = "Year",
     y = "CO2 Emissions", col = "Region") 
       
```


For the most part, the total CO2 emissions of countries have been increasing from 1960 onwards.
Interestingly, Sweden's emissions are decreasing, but it is clear that the emissions for 
several major countries have been rising for decades.





3. (5pt) Now let’s analyze the CO2 emissions per capita (*co2_PC*). Make a similar plot of thesame countries. What does this figure suggest? \

```{r}
gm %>% 
filter( iso2 %in% c("IN","CN","US","SE","VN") ) %>% 
filter(!is.na(co2_PC)) %>% 
ggplot(aes(x=time, y=co2_PC, col=name))+geom_line()+
  labs(x = "Year",
     y = "CO2 Emissions Per Capita", col = "Region") 
       
```

This figure shows that even though America's co2 emissions are increasing as seen in the previous chart,their per capita emissions are decreasing. However, the per capita emissions of several other major countries like India and China are increasing. Sweden remains an outlier as both a major nation and also both lower emissions and per capita emissions.


4. (6pt) Compute average CO2 emissions per capita across the continents (assume *region* is the same as continent). Comment what do you see. \
Note: just compute averages over countries and ignore the fact that countries are of different size. \
Hint: Americas 2016 should be 4.80. \

```{r}
 percont <- gm %>% 
filter(!is.na(co2_PC))%>% 
  filter(time==1960 | time==2016) %>% 
  group_by(region, time) %>% 
  summarise(co2PC=mean(co2_PC))
  percont
       
```
The co2 per capita emissions have increased for every continent from 1960-2016,
with the exception of America. Maybe this is because of more efficient industrial practices
that have developed over time.

5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016. 
Hint: it should look something along these lines(**Please refer to the spec for the graph**) \
```{r}
ggplot(data=percont) %>% +
geom_col(mapping = aes(x=region, y = co2PC, fill=factor(time)),position="dodge")+
  labs(x = "Region", y = "CO2 Emissions Per Capita",  fill= "Year")
#+ scale_fill_discrete(name="Year")
```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume *region* is continent). \

```{r}
#2019 data is unworkable because of excess NA data, used 2016 because it has better data
#the thee largest co2 emitters for each continent:
gm %>%
group_by(region) %>% 
filter(time==2016)%>% 
  filter(!is.na(region),!is.na(name),!is.na(co2_PC)) %>% 
  select(name,co2_PC)%>% 
  filter(rank(desc(co2_PC))<4) %>% 
  arrange(region,co2_PC)


```
The largest co2 emitters involve very rich nations like Qatar, Libya, and Bahrain. For
instance, Libya has rich oil reserves and Qatar has some of the largest natural gas reserves in the world.
```{r}
#the thee smallest co2 emitters for each continent:
gm %>%
group_by(region) %>% 
filter(time==2016)%>% 
  filter(!is.na(region),!is.na(name),!is.na(co2_PC)) %>% 
  select(name,co2_PC)%>% 
  filter(rank(co2_PC)<4) %>% 
  arrange(region,co2_PC)
```
The smallest co2 emitters include poorer nations like Yemen, Somalia, and Haiti. For instance,
while Yemen has rich oil reserves, it has a chaotic government and has been disrupted by civil
war.
---------------------------------------------------------------------

# 4 GDP per capita (50pt)
Let’s look at GDP per capita (*GDP_PC*)

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better. \
Comment what do you see there. \
```{r}
 gm %>% 
 filter(time==1960) %>%  
ggplot(
       aes(GDP_PC, lifeExpectancy,
           col=region, size=totalPopulation/1000000)) +
                           
   geom_point(alpha=0.5) +
   labs(x = "GDP per capita ",
        y = "Life Expectancy", col = "Region",size= "Population\n(Millions)" ) 
   #theme(text = element_text(size, region=15))
```
Higher GDP per capita may be correlated with higher life expectancy. For example, some countries in Europe have very high GDP per capita, passing 20000, and also have life expectancy over 70 years.


2. (4pt) Make a similar plot, but this time use 2019 data only. \

```{r}
 gm %>% 
 filter(time==2019) %>%  
ggplot(
       aes(GDP_PC, lifeExpectancy,
           col=region, size=totalPopulation/1000000)) +
                           
   geom_point(alpha=0.5) +
   labs(x = "GDP per capita ",
        y = "Life Expectancy", col = "Continent",size= "Population\n(Millions)" ) 
   
```
Here we once again see that there is a direct relationship between increased life expectancy
and higher GDP, as the countries with a higher per capita GDP are higher up in terms of lifespan.

3. (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years? \
The GDP per capita and life expectancy has increased for almost every country overall. In 1960, poor, highly populated countries had GDP per capitas under 5000 and life expectancies were around 35-50 years. In 2019, poor and highly populated countries were still underdeveloped compared to more affluent nations, but had GDP per capitas closer to 5000-15000 and life expectancies of around 60 to 75 years. Furthermore, the same countries that were "behind" developmentally in 1960 are mostly
still behind, while richer nations have continued to increase in GDP per capita and life expectancy. Some countries have deviated from this, however, such as the few Asian countries with high GDP
per capita and lifespans comparable to some countries in Europe and the Americas.

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures? \
Note: here as *average*I mean just average over countries, ignore the fact that countries are of different size.\

```{r}
gm %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  filter(time==1960 | time==2016) %>% 
  group_by(region, time) %>% 
summarize(LE=mean(lifeExpectancy))

```
Yes, this matches what I see on the figures: the life expectancy for all countries has increased, and while Asia has caught up with the pace of other countries, Africa still has not, despite some improvements.

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see. \
Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See [**https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute**](https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute).\

```{r}
gm %>% 
filter(!is.na(time),
       !is.na(lifeExpectancy),
       !is.na(region),
       time==1960|time==2016) %>% 
group_by(time,region) %>% 
  summarise(avg=mean(lifeExpectancy))%>% 
  group_by(region) %>% 
  mutate(previous= lag(avg), growth= avg-previous) %>% 
  filter(!is.na(growth))
  
  
```
Interestinly, while Europe has the highest average life expectancy, it has had the least growth
in life expectancy. Conversely, Africa has had a very high growth in life expecctancy comparable
to Asia, but because life expectancy in Africa was very low to begin with, the current life
expectancy is still shorter than other regions.

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it! \
```{r}
gm %>% 
filter(time== 1960 |time==2019) %>% 
filter(!is.na(name), !is.na(region), !is.na(GDP_PC)) %>% 
ggplot(aes(x= GDP_PC, fill=factor(time)))+
geom_histogram( position= "dodge")+
  #theme(axis.text.x=element_text(angle=90, vjust=0.5, injust=1 ))

  labs(x = "GDP Per Capita", y = "Countries",  fill= "Year")
```
GDP per capita overall has increased drastically, with the highest per capita GDP being just over 30000 in 1960 to over 105000 in 2019.


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.) \
Hint: check out the function rank()! \
Hint2: 17 for 1960. \
```{r}
new60 <- gm %>% 
filter(!is.na(lifeExpectancy),!is.na(name),time==1960|time==2019) %>% 
group_by(time) %>% 
mutate(numCountries=n_distinct(name)) %>% 
mutate(rank=rank(desc(lifeExpectancy))) %>% 
filter(name=="United States of America") %>% 
select(name,time,rank) 
new60

```
The US life expectancy ranking has drastically fallen betwen 1960 and 2019.

8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year? \
Hint: 0.0904 for 1960. \
```{r}
gm %>% 
  filter(time==2019|time==1960) %>% 
  group_by(time) %>% 
filter(!is.na(name), !is.na(lifeExpectancy)) %>% 
mutate(rank=rank(desc(lifeExpectancy))  /n_distinct(name)) %>% 
mutate(countries= n_distinct(name))%>% 
filter(name=="United States of America") %>% 
select(name, rank,lifeExpectancy)

```
The relative life expectancy rank has gone from 0.0904 in 1960 to 0.2347 in 2019.


Finally tell us how many hours did you spend on this PS.
```{r}
#around 7 hours
```

